---
title: "Redox classes - why is a differentiation important"
author: "Kristof Dorau"
date: '2021-02-09'
slug: null
draft: true
categories: []
tags:
- Introduction
- Interpretation
subtitle: ''
summary: 'Some words about the concept of redox classes'
authors: []
featured: no
image:
  caption: 
  focal_point: Smart
  preview_only: no
projects: []
---

*The post was last updated on `r format(Sys.Date(), "%Y-%m-%d")`.*

<style>
p.caption {
  font-size: 0.6em;
}
</style>


```{r echo = FALSE, message = FALSE}

knitr::opts_chunk$set(warning = F, echo = F, error = F, message = F)

library(tidyverse)
library(widgetframe)
library(timevis)
library(rmarkdown)
library(kableExtra)

## The theme must be in FRONT (!) of all ggplot or it will be overwritten
my_theme <- theme(
  panel.background = element_rect(fill = "white", color = "black"),
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.border = element_blank())


```

The determination of the soil's redox status is important and impacts a variety of processes of environmental concern:

* greenhouse gas emission from soils,
* nutrient availability,
* pollutant dynamics, and
* pedogenesis are only some examples. 

From a practical point of view, it is not sufficient to state that a soil is reducing or not. It is encouraged to classify and determine redox classes with a more distinct differentiation. 

```{r echo = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)
library(widgetframe)
library(timevis)
library(patchwork)
library(rmarkdown)
library(kableExtra)
library(forcats)
library(ggrepel)

## The theme must be in FRONT (!) of all ggplot or it will be overwritten
my_theme <- theme(
  panel.background = element_rect(fill = "white", color = "black"),
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  panel.border = element_blank())

```



![*Differentiation of redox classes taken from Dorau 2016. The indices refer to the following references: ^1^Yu  and  Patrick  2004; ^2^Shrestha et al. 2014;  ^3^Rennert and  Mansfeldt 2005;  ^4^Matern and  Mansfeldt 2015; ^5^Hindersmann and Mansfeldt 2014; ^6^Dalkmann et al. 2013; ^7^Schuth et al. 2015; ^8^Mansfeldt and Overesch 2013; ^9^Peretyazhko and Sposito 2005; ^10^Schieber 2011; ^11^Morse et al. 1999; ^12^Picek et al. 2000*](01.PNG)

As outlined before, oxygen, manganese, iron, and sulfate are redox sensitive compounds in the soils environment. The transformation (*reduction*) is coupled to the oxidation of soil organic matter and occurs in a predictable sequence. 

> The sequential reduction sequence is synonymously called reddox ladder.

```{r echo = FALSE, message=FALSE}

classes <- readr::read_delim("classes.csv", delim = ";")

options(knitr.kable.NA = '')

classes %>%
  select(Reference, Label, Upper, Lower) %>%
  kbl(caption = "Prominent redox classes in soils at pH 7") %>%
  kable_paper("hover", full_width = T)


#classes
  #filter(author == "Reddy and DeLaune (2008)") %>%
  #mutate(asd = factor(asd = levels(c("one", "two", "three", "four")))) %>%
  #pivot_longer(cols = c("upper", "lower")) %>%
  #ggplot(aes(author, value, color = asd, group = author)) +
  #geom_line(size = 30) +
  #geom_point() +
  #facet_wrap(~author, scales = "free_x")

```

Two things are evident from the table above:

1. The authors label their classes differently (e.g. strongly and highly reducing is somewhat synonym between Zhi-Guang and Reddy & DeLaune) and there is also some overlap between the classes. 
2. Mitsch and Gosselink (2015) do no label the class but link it directly to the dissapearance of oxygen at 250 mV, which is where the transition from nitrogen within nitrate to reduced nitrogen species (e.g. in N~2~O) occurs.

> The E~H~ should not be considered as a **treshold** ❗️ because there are other factors such as temperature and pH which impact transformation rates in soils. 

<style>
p.caption {
  font-size: 0.6em;
}
</style>

